A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-03 latex/main.tex
Updates from Overleaf

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
leading to XA, Xx 
(the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x.}
\label{fig:averted_mutations}
\end{figure}

@@

\subsubsection*{Likely spurious influential recombinants}
In the sc2ts ARG, there are two recombination events which have a large number of
descendants that are not classified as recombinants by Pango. The first of these
corresponds to recombination event within BA.2 whose breakpoint is on the far right
(position 27382) supported by only 4 sites / 3 loci, and which therefore does not
pass our recombinant quality control measure. The second of these involves a single
breakpoint at corresponds to
the origin of the BA.5 lineage

@@

We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
as described in \ref{sec:supplement_pango_lineages}, and
omit Pango designations starting with "X",
to avoid recombinants, leaving 1329 Pango lineages.
To ensure that each sample is a tip (rather than an internal node)
in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 lineages
which are singleton descendants of a recombination event in the ARG
(representing potentially missed Pango recombinant lineages),
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an "X".
This results in a simplified ARG of 1282 samples,
each sample having a different Pango designation.

The simplified ARG contains two single-breakpoint recombination nodes. The first
 The second is the origin of BA.5 which we also suggest may be artefactual (see Supp FIXME).
To reduce plot density, we consider only those Pango lineages with fewer than 10 samples,
and extract non-recombinant structures from the ARG
by creating a ``base tree'' of 535 Pangos (i.e. excluding all descendants of Delta, BA.2, and BA.5),
a Delta subtree of 160 Pangos,
a BA.2 subtree of 151 sample (excluding BA.5 descendants) and
a BA.5 subtree of 251 samples.
Figure SX shows each of these trees compared against the equivalent dated UShER topology,
with the left and right parents of Delta and BA.2 highlighted in the base tree.
The order of tips in the two trees was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
The entire pipeline is saved as a notebook at URL (?).

The majority of resolvable nodes, including most high-level Scorpio groupings,
are shared between the ARG and the UShER tree (magenta nodes).
In the few cases where the trees differ, this is often in terms of degree of resolution of polytomies.
In these cases, sometimes the sc2ts ARG shows greater resolution,
as in the base of the BA.2 subtree, and
sometimes the UShER tree shows greater resolution, as in BA.2.9.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
Mapping deletions onto the ARG using post-hoc parsimony
allowed us to identify important deletions and
to clarify their nature in SARS-CoV-2 evolution.
In Table~\ref{tab:major_dels},
we list a subset of deletion events in the ARG
which are inherited by at least 10,000 samples.
We consider these deletions to be major events
in the evolutionary history of SARS-CoV-2.
Most of these deletion events
(besides the 1-base deletion at 28,271
inherited by 45\% of the samples)
are associated with the origin nodes of major VOCs in the ARG
(indicated by the Pango lineages assigned to them,
e.g., node 86456 is the Alpha origin node;
also see Figure~\ref{fig:subgraph_alpha}).
Most of them are characteristic mutations
that are linked to the emergence of these VOCs
(see the references in Table~\ref{tab:major_dels}
pointing to the analyses
which originally identified them as characteristic mutations).
They are also highly prevalent
(all except two events in this table
are inherited by at least 11\% of the samples).
Notably, eight of these deletion events occur
in the N-terminal domain of the S1 unit of Spike (S1-NTD),
a known hotspot of recurrent deletions \cite{McCarthy2021};
These are all highly recurrent (Table~\ref{tab:rec_dels}), and
include known recurrent deletions, for example,
a 6-base deletion at 21,765 \cite{Meng2021} and
a 3-base deletion at 21,991 \cite{McCarthy2021}.

Detection of deletions and inference of deletion events
remain challenging, especially when the dataset is enormous.
Here we took a naive approach to map deletions onto the ARG,
treating each site involved in a deletion
as independent of the other sites.
Consequently, some deletion events might be split up and
appear as multiple smaller deletion events,
which are likely artefactual.
The three contiguous out-of-frame deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
are examples of such artefacts (Table~\ref{tab:major_dels}).
In the sequence alignments, there are two 9-base deletions,
one at 11,283 and the other at 11,288,
which are commonly found in BA.1 and BA.2 samples, respectively.
Assuming that the alignments are correct,
there should be two deletion events,
but they were mapped as three deletion events
as they had four overlapping sites from 11,288 to 11,292,
which were inferred by parsimony
as the 4-base deletion event at 11,288.
Furthermore, sequencing errors or sequence misalignments
can also cause artefactual deletions, for example,
the out-of-frame 4-base recurrent deletion at 21,767 in S1-NTD
(Table~\ref{tab:rec_dels}).
Future improvements in deletion remapping and data pre-processing
should remove such artefactual deletions.

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
The deletions caused by these events
are inherited by at least 10,000 samples.
The frequencies of these deletions in the ARG are shown.
The Pango lineage assigned to the node
where an event occurs is shown.
The references point to analyses
which identified a deletion
to be associated with a major VOC assigned to its node.
Abbreviations: NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:major_dels}
\begin{tabular}{rccrcrrc}
\toprule
Start & Length & Region & Node & Pango lineage & Frequency (\%) & Ref\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 45.06 &  \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 44.91 & \cite{Stern2021} \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 44.82 & \cite{Stern2021} \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 35.12 &  \\
28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 35.12 & \cite{RambautOmicron2021} \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 21.44 & \cite{RambautOmicron2021} \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 21.43 &  \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 13.73 & \cite{RambautOmicron2021} \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 13.69 &  \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 13.68 & \cite{RambautOmicron2021} \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 13.68 & \cite{RambautOmicron2021} \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 12.60 & \cite{RambautOmicron2021} \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 11.70 &  \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 11.70 & \cite{RambautAlpha2020} \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 11.64 & \cite{RambautAlpha2020} \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 11.70 & \cite{RambautAlpha2020} \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 7.67 & \cite{Tegally2022} \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 0.94 &  \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{
Recurrent deletions.
These deletions occur at least 5 times in the ARG and
are found in at least 1,000 samples.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:rec_dels}
\begin{tabular}{rrcrr}
\toprule
Start & Length & Region & Occurrences & Samples\\
\midrule
686 & 9 & ORF1ab / NSP1 & 6436 & 43354\\
21991 & 3 & S / S1-NTD & 2088 & 309629\\
21765 & 6 & S / S1-NTD & 482 & 820194\\
22029 & 6 & S / S1-NTD & 275 & 1107941\\
22194 & 3 & S / S1-NTD & 200 & 310349\\
6513 & 3 & ORF1ab / NSP3 & 181 & 341380\\
21987 & 9 & S / S1-NTD & 70 & 338633\\
28248 & 6 & ORF8 & 28 & 1113392\\
28271 & 1 & non-coding & 24 & 1331409\\
11288 & 9 & ORF1ab / NSP6 & 19 & 843291\\
11283 & 9 & ORF1ab / NSP6 & 13 & 339810\\
21767 & 4 & S / S1-NTD & 10 & 2144\\
28362 & 9 & N & 7 & 850483\\
21633 & 9 & S / S1-NTD & 7 & 531973\\
\bottomrule
\end{tabular}
\end{table}

@@

% https://github.com/jeromekelleher/sc2ts-paper/issues/772
\subsubsection{Alpha}
The B.1.1.7 (Alpha) variant is the first major VOC, and
it is a saltational lineage
characterized by a long branch preceding its origin.
Figure \ref{fig:subgraph_alpha} shows that the origin of Alpha is well reconstructed in the ARG.
Most of the characteristic Alpha mutations (\cite{RambautAlpha2020})
(20 out of 23) in the ARG
occur immediately above the Alpha root.
However, a characteristic three-base correlated mutation
is instead associated with the parent node (labeled B.1.1) of the Alpha root,
suggesting that this mutation was acquired prior to the emergence of Alpha.
This parent node has an additional 5 children
(leading to 12 samples),
of which we display one (ERR4413600)
with no associated mutations, meaning that it has an identical sequence to the inferred ancestor of Alpha.
Below the root of Alpha, a single sample (ERR5178290)
appears without the characteristic mutation C3267T
(or any additional mutations).
However, this is only a single sample and
given that it also appears two months after the first Alpha samples
we treat its placement with caution.
Note that the list of mutations in \cite{RambautAlpha2020}
includes a T26801C in the M gene
which should no longer be considered characteristic of Alpha,
as the Wuhan-Hu-1 reference at 26,801 has since been updated to C.
In the ARG, a one-base non-coding deletion (position 28,271)
is also associated with the Alpha root;
although this is not in the original list of
characteristic Alpha mutations,
it has been suggested elsewhere as an Alpha-defining mutation 
(\cite{Li2023indels}).

\subsubsection{Delta}
 \ref{fig:subgraph_delta}

\subsubsection{Omicron BA.1 and BA.2}
XX

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{recombinant_qc}
\caption{
Quality control of recombination events.
Scatterplot of recombination events by the net number of supporting loci
on the left parent (x-axis) and the right parent (y-axis).
The shaded regions highlight artifactual recombination events,
which have fewer than 4 net supporting loci
on one side or both sides of the suggested breakpoint.
Colors indicate the primer scheme used for sequencing.
The pie charts show breakdowns of the recombination events
by primer scheme per quadrant (labeled Q1 to Q4).
Recombination events associated with
the origins of Pango X lineages are labeled.
}
\label{fig:recombinant_qc}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Delta.pdf}
\caption{
Subgraph illustrating the saltational origin of B.1.617.2 (Delta). Symbols as in \ref{fig:subgraph_alpha}, with characteristic Delta mutations outlined in pink. Sites associated with multiple mutations in the subgraph have their mutations assigned a unique fill colour (orange, blue, green, red, etc); reversions are further highlighted with a black outline. Sc2ts seed samples (see STAR methods) for B.1.617, B.1.617.1, and B.1.617.2 (Delta) are plotted as red squares.
}
\label{fig:subgraph_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Omicron.pdf}
\caption{
Subgraph of the BA.1 and BA.2 Omicron lineages. BA.1 samples are shown 
[Some notes to help the reader know this 
reflects known events]
}
\label{fig:subgraph_omicron}
\end{figure}

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson group B (JB) and Jackson singleton recombinant 1 (J1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns (STAR Methods, Figure~\ref{fig:copying_pattern_examples}) 
leading to XA, XW, XBB, and an
undesignated recombination event under XBB, which we here refer to as  XBB.x.}
\label{fig:averted_mutations}
\end{figure}

@@

\subsubsection*{Likely spurious ancestrally-important recombinants}
\label{sec:spurious-ancestrally-important-recombinants}
In the sc2ts ARG, there are two recombination events which have a large number of
descendants but which are not classified as recombinants by Pango. The first of these
corresponds to recombination event within BA.2 whose breakpoint is at position 27382
on the far right of the genome (see supplementary document S3, node 1030562). Although
supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
meaning that this event does not pass our recombinant quality control measure.

The second of these involves a single breakpoint at position 26858, and is the origin
of the BA.5 lineage in the ARG (see supplementary document S3, node 1189192). Although
supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
sample is added to the ARG, it is possible to construct a more parsimonious
non-recombinant origin by long branch splitting (see the \nameref{sec:arg_postprocessing}
section), although subsequent tree building does not make this a parsimonious
way to rewire the final ARG. We therefore suggest that this too could be an artifactual
recombination event.

@@

We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
as described in \ref{sec:supplement_pango_lineages}, and
omit Pango designations starting with "X"
to avoid recombinants, leaving 1329 Pango lineages.
To ensure that each sample is a tip (rather than an internal node)
in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 lineages
which are singleton descendants of a recombination event in the ARG
(representing potentially missed Pango recombinant lineages),
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an "X".
This results in a simplified ARG of 1282 samples,
each sample having a different Pango designation.

The simplified ARG contains only two recombination nodes, which
correspond to those on the far right of the genome identified as
artefactual in the \ref{sec:spurious-ancestrally-important-recombinants} section.
We therefore compare the Usher topology against the first tree in the ARG (covering
positions 1-26858; the majority of the genome). To reduce plot density, we consider
only those Pango lineages with greater than 10 samples, and to aid visualization,
we divide resulting tree into 4 parts: a BA.5 subtree of 187 Pangos,
a Delta subtree of 104 Pangos, a BA.2 subtree (excluding BA.5 descendants) of 118 Pangos,
and a base tree (i.e. excluding all descendants of Delta and BA.2) of 405 Pangos,
Figures \ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} show these trees compared
against the equivalent dated UShER topologies.
The order of tips in each plot was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
See \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/supp\_pdf-cophylogeny-tanglegrams.ipynb}{supp\_pdf-cophylogeny-tanglegrams.ipynb}
for details and the pipeline used.

Inspection of Figures~\ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} reveals that the majority of resolvable nodes,
including most high-level Scorpio groupings,
are shared between the ARG and the UShER tree (magenta nodes), and there is extremely close (usually perfect) agreement on topological relations between the Pango lineages.
In the few cases where the trees differ, this is often in terms of degree of resolution of polytomies.
In these cases, sometimes the sc2ts ARG shows greater resolution,
as in the base of the BA.2 subtree, and
sometimes the UShER tree shows greater resolution, as in the close relatives of Beta (B.1.351).

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
Although uncommon relative to single nucleotide mutations,
deletions are an important component of SARS-CoV-2 evolution,
contributing to the emergence of VOCs by enhancing,
e.g., efficiency for cell entry \cite{Cantoni2022} and
potential for antibody escape \cite{McCarthy2021}.
Using the ARG, we identified major deletion events
(which are propagated to many descendants)
as well as recurrent deletions
(which occur on at least two branches) in the ARG.
In Table~\ref{tab:major_dels},
we list the major deletion events in the ARG
which are each inherited by at least 10,000 descendant samples; and
in Table~\ref{tab:rec_dels},
we list the recurrent deletions in the ARG
which are each observed by at least 10,000 (0.4\%) samples.
Most of the deletions in these tables
occur in coding regions
(except the 1-base deletions at 28,271) and
do not cause frame shifts, involving 3, 6, or 9 sites
(except the 4-base deletion and the two 5-base deletions,
which are likely artefacts; see below).
All the deletions in these tables
(besides the artefactual cases)
have been linked to the emergence of major VOCs
(Alpha \cite{RambautAlpha2020},
Delta \cite{Stern2021},
Omicron B.1.1.529 \cite{RambautOmicron2021},
BA.1 \cite{RambautOmicron2021},
BA.2 \cite{RambautOmicron2021}, and
BA.4 \cite{Tegally2022})
and/or reported to be prevalent
in previous studies of deletions \cite{Li2023indels}.
All these major deletion events
(again, besides the artefactual cases)
are associated with the origins of major VOCs in the ARG
(see the Alpha subgraph in Figure~\ref{fig:subgraph_alpha};
the Delta subgraph in Figure~\ref{fig:subgraph_delta}; and
the Omicron subgraph in Figure~\ref{fig:subgraph_omicron}),
except the 1-base deletion at 28,271 on node 1436808
and the 9-base deletion at 686 on the BA.4-labeled node 2698748
(Table~\ref{tab:major_dels}).
Most of these deletions have been identified
as characteristic mutations of the major VOCs
in the studies or analysis cited in Table~\ref{tab:major_dels}.
Notably, eight of the major deletion events occur
in the N-terminal domain of the S1 unit of Spike
(a known hotspot of recurrent deletions \cite{McCarthy2021}
that plays a substantial role in antigenicity \cite{Harvey2021}),
which are all highly recurrent in the ARG
(Table~\ref{tab:rec_dels}).
These include known recurrent deletions:
the 6-base deletion at 21,765 and
the 3-base deletion at 21,991,
which are associated with
increased efficiency in cell entry \cite{Meng2021} and
antibody escape \cite{McCarthy2021}, respectively.
In addition, we found recurrent deletions in ORF8 and NSP3,
which have been suggested to be indel hotspots
\cite{Akaishi2022}.
Together, these findings demonstrate that
the ARG captures genuine deletion events
in SARS-CoV-2 evolution.
A comprehensive evolutionary analysis of
the deletion events in the ARG
is beyond the scope of the current study;
however, we believe that the ARG, as a resource,
will enable and facilitate future studies
to better understand the role of deletions
in SARS-CoV-2 evolution
(alongside recombination and single nucleotide mutations),
which to-date has been understudied
in great part due to
a lack of suitable data models, methods, and tools.

Detection of deletions and inference of deletion events
remain challenging, especially when the dataset is enormous.
Here we took a naive approach to map deletions onto the ARG,
treating each site involved in a deletion
as independent of the other sites.
Consequently, some deletion events might be split up and
appear as multiple smaller deletion events,
which are likely artefactual.
The three contiguous out-of-frame deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
are examples of such artefacts (Table~\ref{tab:major_dels}).
In the sequence alignments, there are two 9-base deletions,
one at 11,283 and the other at 11,288,
which are found in
339,810 BA.1 samples and 843,291 BA.2 samples
in the Viridian v04 dataset, respectively.
Assuming that the alignments are correct,
there should be two deletion events,
but they were mapped as three deletion events
as they had four overlapping sites from 11,288 to 11,292,
which were inferred by parsimony
as the 4-base deletion event at 11,288.
Future improvements in deletion remapping
should remove such artefactual deletions.

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
The deletions caused by these events
are inherited by at least 10,000 samples.
These deletions are characterized by
a site position in the genome (start) and
the number of subsequence sites involved (length), and
they are annotated by the gene in which they occur
(non-coding otherwise).
The IDs of the nodes in the ARG
above which the deletions occur and
the Pango lineages assigned to these nodes are indicated.
The percentage of samples inheriting each deletion
is shown as frequencies.
If a deletion has been identified
as a characteristic mutation of a VOC by a study (or an analysis),
a reference to the study (or the analysis) is provided.
Abbreviations: NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:major_dels}
\begin{tabular}{rccrcrrc}
\toprule
Start & Length & Region & Node & Pango lineage & Frequency (\%) & Ref\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 45.06 &  \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 44.91 & \cite{Stern2021} \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 44.82 & \cite{Stern2021} \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 35.12 &  \\
28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 35.12 & \cite{RambautOmicron2021} \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 21.44 & \cite{RambautOmicron2021} \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 21.43 &  \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 13.73 & \cite{RambautOmicron2021} \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 13.69 &  \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 13.68 & \cite{RambautOmicron2021} \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 13.68 & \cite{RambautOmicron2021} \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 12.60 & \cite{RambautOmicron2021} \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 11.70 &  \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 11.70 & \cite{RambautAlpha2020} \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 11.64 & \cite{RambautAlpha2020} \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 11.70 & \cite{RambautAlpha2020} \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 7.67 & \cite{Tegally2022} \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 0.94 &  \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{
Recurrent deletions in the ARG.
These deletions occur on at least two branches in the ARG and
are found in the alignments of at least 10,000 (0.4\%) samples
in the Viridian v04 dataset.
For each deletion, the number of occurrences in the ARG and
the number of samples where it is observed are shown.
For further explanation,
see the caption of Table~\ref{tab:major_dels}.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:rec_dels}
\begin{tabular}{rrcrr}
\toprule
Start & Length & Region & Occurrences & Samples\\
\midrule
686 & 9 & ORF1ab / NSP1 & 6436 & 43354\\
21991 & 3 & S / S1-NTD & 2088 & 309629\\
21765 & 6 & S / S1-NTD & 482 & 820194\\
22029 & 6 & S / S1-NTD & 275 & 1107941\\
22194 & 3 & S / S1-NTD & 200 & 310349\\
6513 & 3 & ORF1ab / NSP3 & 181 & 341380\\
21987 & 9 & S / S1-NTD & 70 & 338633\\
28248 & 6 & ORF8 & 28 & 1113392\\
28271 & 1 & non-coding & 24 & 1331409\\
11288 & 9 & ORF1ab / NSP6 & 19 & 843291\\
11283 & 9 & ORF1ab / NSP6 & 13 & 339810\\
28362 & 9 & N & 7 & 850483\\
21633 & 9 & S / S1-NTD & 7 & 531973\\
\bottomrule
\end{tabular}
\end{table}

@@

% https://github.com/jeromekelleher/sc2ts-paper/issues/772
\subsubsection{Alpha}
\label{supp:alpha}
The B.1.1.7 (Alpha) variant emerged in the UK as the first major VOC. Is well reconstructed in the ARG with Figure \ref{fig:subgraph_alpha} showing it to be a saltational lineage characterised by a long branch preceding its origin. 
Most of the characteristic Alpha mutations (\cite{RambautAlpha2020})
(20 out of 23) in the ARG
occur immediately above the Alpha root.
However, a characteristic three-base correlated mutation
is instead associated with the parent node (labeled B.1.1) of the Alpha root,
suggesting that this mutation was acquired prior to the emergence of Alpha.
This parent node has an additional 5 children
(leading to 12 samples),
of which we display one (ERR4413600)
with no associated mutations, meaning that it has an identical sequence to the inferred ancestor of Alpha.
Below the root of Alpha, a single sample (ERR5178290)
appears without the characteristic mutation C3267T
(or any additional mutations).
However, this is only a single sample and
given that it also appears two months after the first Alpha samples
we treat its placement with caution.
Note that the list of mutations in \cite{RambautAlpha2020}
includes a T26801C in the M gene
which should no longer be considered characteristic of Alpha,
as the Wuhan-Hu-1 reference at 26,801 has since been updated to C.
In the ARG, a one-base non-coding deletion (position 28,271)
is also associated with the Alpha root;
although this is not in the original list of
characteristic Alpha mutations,
it has been suggested elsewhere as an Alpha-defining mutation 
(\cite{Li2023indels}).

\subsubsection{Delta}
\label{supp:delta}

 \ref{fig:subgraph_delta}

\subsubsection{Omicron BA.1 and BA.2}
\label{supp:omicron}

\ref{fig:subgraph_omicron}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{recombinant_qc}
\caption{
Quality control of recombination events.
Scatterplot of recombination events by the net number of supporting loci
on the left parent (x-axis) and the right parent (y-axis).
The shaded regions highlight potentially artifactual recombination events,
which have fewer than 4 net supporting loci
on one side or both sides of the suggested breakpoint.
Colors indicate the primer scheme used for sequencing.
The pie charts show breakdowns of the recombination events
by primer scheme per quadrant (labeled Q1 to Q4).
Recombination events associated with
the origins of Pango X and Jackson\cite{Jackson2021} lineages are labeled.
}
\label{fig:recombinant_qc}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Delta.pdf}
\caption{
Subgraph illustrating the saltational origin of B.1.617.2 (Delta). Symbols as in \ref{fig:subgraph_alpha}, with characteristic Delta mutations outlined in pink. Sites associated with multiple mutations in the subgraph have their mutations assigned a unique fill colour (orange, blue, green, red, etc); reversions are further highlighted with a black outline. Sc2ts seed samples (see STAR methods) for B.1.617, B.1.617.1, and B.1.617.2 are plotted as red squares.
}
\label{fig:subgraph_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Omicron.pdf}
\caption{
Subgraph illustrating the saltational origin of the major Omicron lineages BA.1 and BA.2. Symbol and multiple-colouring schemes as in \ref{fig:subgraph_delta}, except BA.1 samples are filled in light blue (also used to outline BA.1 specific mutations), BA.2 samples in light yellow (also used to outline BA.2 specific mutations), and B.1.1.529 mutations not seen in either BA.1 or BA.1 are outlined in green. These colours correspond to those used in the Pango designation issue\cite{RambautOmicron2021}.
}
\label{fig:subgraph_omicron}
\end{figure}
